---
title: "Individualized Explainable Artificial Intelligence: A Tutorial For Identifying Local and individual Predictions"
execute:
  message: FALSE
author: 
   - name: "Mohammed Saqr"
     email: "mohammed.saqr@uef.fi"
     affil-id: 1,*
   - name: "Sonsoles López-Pernas"
     email: "sonsoles.lopez@uef.fi"
     affil-id: 1
affiliations:
  - id: 1
    name: "University of Eastern Finland"
  - id: "*"
    name: "Corresponding author: Mohammed Saqr, `mohammed.saqr@uef.fi`"
---


```{r, include=FALSE}
# Register an inline hook:
knitr::knit_hooks$set(inline = function(x) {
  x <- sprintf("%1.3f", x)
  paste(x, collapse = ", ")
})
```

# Local Explanations


```{r setup, include=FALSE}
library(tidyverse)
library(DALEX)
library(DALEXtra)
library(tidymodels)
library(e1071)
library(rio)
student_data <- import("https://github.com/lamethods/data2/raw/main/lms/lms.csv")
options(scipen = 999)
```

## Building a machine learning model for explanation


```{r}
# Set a seed for reproducibility of results
set.seed(265)
```


```{r}
# Center the selected predictor variables in the dataset. This involves 
# subtracting the mean from each value, making the mean of each variable zero
student_data_centered <- student_data |>
  mutate(across(Freq_Course_View:Active_Days, ~scale(., scale = FALSE) |> as.vector()))
```


```{r}
# Split the centered dataset into training (80%) and testing (20%) sets
data_split <- initial_split(student_data_centered, prop = 0.8)
train_data <- training(data_split)  # Extract the training set
test_data <- testing(data_split)    # Extract the testing set
```


```{r}
# Define the formula specifying the relationship between the target variable
#  (Final_Grade) and the predictors (Freq_Course_View, Freq_Lecture_View, etc.)
formula <- Final_Grade ~ Freq_Course_View + Freq_Lecture_View + Freq_Forum_Consume +
  Freq_Forum_Contribute + Regularity_Course_View +
  Regularity_Lecture_View + Regularity_Forum_Consume +
  Regularity_Forum_Contribute + Session_Count +
  Total_Duration + Active_Days

# Fit a Support Vector Machine (SVM) model using the radial kernel
# The model is trained on the training dataset with the specified formula
svm_fit_radial <- svm(formula, data = train_data, kernel = "radial")
```



```{r}
# Create an explainer for the SVM model using the DALEX package
explainer_svm <- explain(
  model = svm_fit_radial,    # The trained SVM model with a radial kernel
  data = test_data[,-12],    # Test data excluding the target variable (Final_Grade)
  y = test_data$Final_Grade, # True values of the target variable for comparison
  label = "SVM Radial"       # Label for the model in the explainer
)

# Check the explainer
print(explainer_svm)
```

## Breakdown profiles




### Creating and visualizing breakdown results through waterfall plots


```{r}
Instance5 <- test_data[5, ]  # Select any other observation from `test_data` if needed
```


```{r}
# Set a seed for reproducibility
set.seed(265)

breakdown_explanation_5 <- predict_parts(
  explainer = explainer_svm,
  new_observation = Instance5,
  type = "break_down"
)
```


```{r}
print(breakdown_explanation_5) # Print the breakdown explanation to the console
```


```{r }
# Plot the breakdown explanation to visualize the contribution of each feature
plot(breakdown_explanation_5)
```

 

### Order of variables


```{r, results = FALSE}
set.seed(265)

# Define a custom order for the features in the breakdown analysis
Custom_order <- c("Regularity_Course_View", "Regularity_Lecture_View", 
                  "Regularity_Forum_Consume", "Regularity_Forum_Contribute", 
                  "Session_Count", "Total_Duration", "Active_Days", 
                  "Freq_Course_View", "Freq_Lecture_View",
                  "Freq_Forum_Consume", "Freq_Forum_Contribute")

# Select the fifth instance from the test data for local explanation
Instance5 <- test_data[5, ]

# Generate a breakdown explanation using the custom order of features
breakdown_explanation_5_ordered <- predict_parts(
  explainer = explainer_svm,           # The model explainer
  new_observation = Instance5,         # The instance to explain
  order = Custom_order,                # Apply the custom feature order
  type = "break_down"                  # Use the breakdown method
)

# Display and plot the breakdown explanation
print(breakdown_explanation_5_ordered)
```

 
```{r}
plot(breakdown_explanation_5_ordered)
```
 


```{r fig.height=18, fig.width=24}
#| label: fig-breakdown-all
#| fig-cap: Breakdown explanation to visualize the contribution of each feature for different feature orders
set.seed(265)

library(patchwork)  # For combining plots

# Generate 12 different plots with different feature orders
plots <- list()
for (i in 1:length(Custom_order)) {
  random_order <- sample(Custom_order)
  # Compute breakdown explanation with the current order
  breakdown_explanation_ordered <- predict_parts(
    explainer = explainer_svm,
    new_observation = Instance5,
    order = random_order,  # Apply the random feature order
    type = "break_down"
  )
  
  # Plot the breakdown explanation
  plots[[i]] <- plot(breakdown_explanation_ordered) +
    ggtitle(paste("Breakdown Explanation - Order", i))
}

# Combine all plots into a single image
combined_plot <- wrap_plots(plots, ncol = 3)  # Adjust ncol as needed
print(combined_plot)
```

## SHAP (Shapley Additive Explanations) Values: A Detailed Explanation


### Advantages of SHAP Values


### Using SHAP for Single Instance Explanation


```{r}
# Compute SHAP explanation for a single instance
# 'type = "shap"' specifies that we want to use SHAP (SHapley Additive 
# exPlanations) to explain the prediction.
shap_explanation <- predict_parts(
  explainer = explainer_svm,    # The explainer object for the SVM model
  new_observation = Instance5,  # The specific instance that we want to explain
  type = "shap"                 # Use the SHAP method for generating the explanation
)

# Print the SHAP explanation
print(shap_explanation) 
```

```{r}
#| label: fig-shap-exp5
#| fig-cap: SHAP Explanation for Instance 5

# Plot the SHAP explanation
plot(shap_explanation) 
```


## LIME (Local Interpretable Model-Agnostic Explanations) explanations


```{r, fig.width=7, fig.height=3.5}
# Ensure the necessary DALEXtra methods are loaded
model_type.dalex_explainer <- model_type.dalex_explainer
predict_model.dalex_explainer <-  predict_model.dalex_explainer

# Generate LIME explanation using DALEXtra's predict_surrogate function
Lime_Explainer <- predict_surrogate(
  explainer = explainer_svm,    # The explainer object for the SVM model
  new_observation = Instance5,  # The specific instance (Instance5) to be explained
  n_features = 12,              # Number of features to include in the explanation
  n_permutations = 1000,        # Number of permutations for the surrogate model
  type = "lime"                 # Specify that we want a LIME explanation
)

# Plot the LIME explanation
plot(Lime_Explainer) 
```


### Multiple instances

```{r}
# Predict the Final_Grade on the test dataset
predictions <- predict(svm_fit_radial, newdata = test_data)

# Calculate residuals (the difference between actual grades and predicted grades)
residuals <- test_data$Final_Grade - predictions

# Define the threshold for significant prediction errors (10 points)
threshold <- 10

# Identify students whose predicted grades were increased by more than the threshold
increased_grades <- test_data |>
  mutate(Residual = residuals) |>
  filter(Residual <= -threshold)

# Identify students whose predicted grades were decreased by more than the threshold
decreased_grades <- test_data |>
  mutate(Residual = residuals) |>
  filter(Residual >= threshold)

```

```{r}
# Aggregate SHAP values for students with increased grades
shap_increased <- predict_parts(
  explainer = explainer_svm,
  new_observation = increased_grades |> dplyr::select(-Final_Grade, -Residual),
  type = "shap",
  B = 50  # Number of Monte Carlo simulations for SHAP values
)

# Plot the aggregated SHAP values for increased grades
plot(shap_increased)  
```

```{r}
# Aggregate SHAP values for students with decreased grades
shap_decreased <- predict_parts(
  explainer = explainer_svm,
  new_observation = decreased_grades |> dplyr::select(-Final_Grade, -Residual),
  type = "shap",
  B = 50  # Number of Monte Carlo simulations for SHAP values
)

# Plot the aggregated SHAP values for decreased grades
plot(shap_decreased)
```

## Explanation of a classifier model

```{r, warning=FALSE}
# Step 1: Center the numeric variables in the dataset
student_data_centered <- student_data |>
  mutate(across(Freq_Course_View:Active_Days, ~scale(., scale = FALSE) |> as.vector()))

# Step 2: Create a binary classification target variable
median_grade <- median(student_data_centered$Final_Grade)
student_data_centered <- student_data_centered |>
  mutate(Achievement = ifelse(Final_Grade < median_grade, 
                              "Low_Achievers", "High_Achievers")) |>
  mutate(Achievement = factor(Achievement, 
                              levels = c("High_Achievers","Low_Achievers")))

# Step 3: Split the data into training and testing, stratifying by the target variable
data_split <- initial_split(student_data_centered, prop = 0.8, strata = Achievement)
train_data <- training(data_split)
test_data <- testing(data_split)

# Step 4: Train an SVM model using e1071
svm_model_Classify <- svm(
  Achievement ~ Freq_Course_View + Freq_Lecture_View + Freq_Forum_Consume +
    Freq_Forum_Contribute + Regularity_Course_View + 
    Regularity_Lecture_View + Regularity_Forum_Consume +
    Regularity_Forum_Contribute + Session_Count + 
    Total_Duration + Active_Days,
  data = train_data,
  probability = TRUE)

# Step 5: Create explainer
explainer_svm_Classify <- explain(
  svm_model_Classify,
  data = test_data |> dplyr::select(-Final_Grade, -Achievement),
  y = test_data$Achievement,
  label = "SVM Classification",
  verbose = FALSE
)
```


```{r, layout = c(1,1)}
# Select a single instance for local explanation
Instance10 <- test_data[10, ]  # You can select any observation from test_data

# Local interpretation using Break Down method
breakdown_Classify <- predict_parts(
  explainer = explainer_svm_Classify,
  new_observation = Instance10 |> dplyr::select(-Final_Grade, -Achievement),
  type = "break_down")

# Display and plot the breakdown explanation
plot(breakdown_Classify)


# Local interpretation using SHAP method
Shap_Classify <- predict_parts(
  explainer = explainer_svm_Classify,
  new_observation = Instance10 |> dplyr::select(-Final_Grade, -Achievement),
  type = "shap")

# Display and plot the SHAP explanation
plot(Shap_Classify)
```

