---
title: "Idiographic Networks: A tutorial on Graphical Vector Autoregression and Unified Structural Equation Modeling"
author: 
   - name: "Mohammed Saqr"
     email: "mohammed.saqr@uef.fi"
     affil-id: 1,*
   - name: "Daryn Dever"
     email: "ddever@ufl.edu"
     affil-id: 2
   - name: "Sonsoles López-Pernas"
     email: "sonsoles.lopez@uef.fi"
     affil-id: 1
affiliations:
  - id: 1
    name: "University of Eastern Finland"
  - id: 2
    name: "University of Florida"
  - id: "*"
    name: "Corresponding author: Mohammed Saqr, `mohammed.saqr@uef.fi`"
---

## Loading data and pre-processing

```{r setup, include=FALSE}
library(tidyverse)        # For data manipulation and visualization
library(skimr)            # For summary statistics of data frames
library(graphicalVAR)     # Assumed necessary for specific analyzes
library(qgraph)           # Plotting networks
library(pompom)           # uSEM implementation
library(caret)            # For identifying near zero variance predictors
library(rio)              # For importing the data
source("aux.R")
```

```{r}
# Step 1: Load the dataset
df <- import("https://github.com/lamethods/data2/raw/main/srl/srl.RDS")

# Step 2: Define variables for the graphical VAR model
Vars <- c("planning", "monitoring", "effort", "control", 
          "help", "social", "organizing", "feedback", "evaluating")

# Step 3: Create a "Day" variable for each entry within a person
df <- df |>
  group_by(name) |>
  mutate(Day = seq_along(name)) |>
  ungroup()

# Step 4: Filter the data for a specific individual named 'Grace'
Grace_Data <- df |>
  filter(name == "Grace") |>
  dplyr::select(all_of(Vars), Day)
```

### Exploring the data

```{r}
Grace_descriptives <- skim(Grace_Data)
print(Grace_descriptives)
```

```{r fig.width=9, fig.height=6}
# Reshape the data to long format
Grace_data_long <- Grace_Data |>
  pivot_longer(cols = Vars, # Assuming 'id' and 'interaction' are the relevant columns
               names_to = "variable", 
               values_to = "value")

# Plot with faceting by variable
ggplot(Grace_data_long, aes(x = Day, y = value)) +
  geom_line(color = "blue") +
  geom_point() +
  geom_smooth(method = "loess", color = "red", se = FALSE) + # Smoothing (LOESS)
  facet_wrap(~variable, scales = "free_y", ncol = 3) +
  theme_minimal() 
```

### Stationarity

```{r}
detrender(Grace_Data, vars = Vars, timevar = "Day")
```

## Estimation of the `graphicalVAR` Model

```{r}
set.seed(265) # for replicability

# Fit the `graphicalVAR` model to the data from 'Grace' with 'lambda_beta' 
# to 0.1 to apply moderate regularization 
GraphVAR_Grace <- graphicalVAR(Grace_Data, beepvar = "Day", lambda_beta = 0.1)

# Output a summary of the fitted model to review the model's structure, 
# including estimated parameters and fit indices.
summary(GraphVAR_Grace)
```
 
## Visualization and interpretation of the `graphicalVAR` Model


```{r fig.height=8, fig.width=8, layout = c(1,1)}
# Visualize the Partial Contemporaneous Correlations (PCC) within the model 
# using a circular layout.
plot(GraphVAR_Grace, include = "PCC", layout = "circle", 
     theme = "colorblind", labels = Vars, titles = F)

# Similarly, visualize the Partial Directed Correlations (PDC), highlighting 
# temporal directed influences between variables.
plot(GraphVAR_Grace, include = "PDC", layout = "circle", 
     theme = "colorblind", labels = Vars, titles = F)
```


```{r fig.height=8, fig.width=8, layout = c(1,1)}
 plotting <- function(Network, title = "", ...){
      qgraph(Network,
       layout = "circle",          # Circular node arrangement
       title = title,              # Setting a title
       loop = 0.7,                 # Curvature for self-loops
       node.width = 2,             # Node width
       repulsion = 0.8,            # Spacing between nodes
       label.font = 1,             # Label font styling
       label.fill.vertical = 1,    
       label.fill.horizontal = 1,  
       esize = 7,                  # Edge thickness
       vsize = 6,                  # Node size
       color = "#ffeece",          # Light pink node color
       edge.label.cex = 1.5,       # Edge label size
       edge.width = 1,             # Edge width
       border.width = 2,           # Border width for nodes
       edge.labels = TRUE,         # Show edge labels
       asize = 3,                  # Arrow size
       labels = Vars,              # Variable labels,
       negDashed = TRUE,           # Negative edges dashed
       theme = "colorblind", ...   )}

plotting(GraphVAR_Grace$PCC)
plotting(GraphVAR_Grace$PDC)
```

## Estimation of multiple idiographic models n\>1

```{r, cache = TRUE}
detrended <- detrender(data = df, vars = Vars, timevar = "Day", groupvar = "name")

# Fit the mlgraphicalVAR model to the data from 'all students with 'lambda_beta' to 0.1 to apply moderate regularization 

GraphVAR_all <- mlGraphicalVAR(detrended, vars = Vars,
                               idvar = "name", lambda_beta = 0.01,
                               subjectNetworks = TRUE, verbose = FALSE)
```

### Plotting the `mlGraphicalVAR` results

```{r fig.height=8, fig.width=8, layout = c(1,1,1)}
plotting(GraphVAR_all$betweenNet, minimum = 0.05)
plotting(GraphVAR_all$fixedPCC, minimum = 0.05)
plotting(GraphVAR_all$fixedPDC, minimum = 0.01)
```

### Plotting and interpreting subject level networks

```{r}
for (i in 20:24) {
  # Apply plotting to subjectPCC and subjectPDC
  plotting(GraphVAR_all$subjectPCC[[i]], minimum = 0.05, 
           title = paste("Contemporaneous network for ", GraphVAR_all$ids[[i]]))
  plotting(GraphVAR_all$subjectPDC[[i]], minimum = 0.05, 
           title = paste("Temporal network for ", GraphVAR_all$ids[[i]]))
}
```

## Unified Structural Equation Modeling (uSEM)

### uSEM analysis

```{r}
Grace_data_centered <- Grace_Data |>
  mutate(across(everything(), ~ scale(.x, center = TRUE, scale = TRUE))) |> 
  select(-Day)
```

```{r}
Usim_Grace <- uSEM(var.number = 9, 
               data = Grace_data_centered, 
               lag.order = 1, 
               verbose = FALSE,
               trim = TRUE)
```

```{r}
Usim_Grace_fit <- model_summary(model.fit = Usim_Grace,
                          var.number = 9, 
                          lag.order = 1)

cfi_val <- sum(Usim_Grace_fit$cfi)
tli_val <- Usim_Grace_fit$tli
rmsea_val <- Usim_Grace_fit$rmsea
srmr_val <- Usim_Grace_fit$srmr

# Print the values of each fit index
cat("CFI:", cfi_val, "->", ifelse(cfi_val > 0.95, "Passed", "Failed"), "\n")
cat("TLI:", tli_val, "->", ifelse(tli_val > 0.95, "Passed", "Failed"), "\n")
cat("RMSEA:", rmsea_val, "->", ifelse(rmsea_val < 0.08, "Passed", "Failed"), "\n")
cat("SRMR:", srmr_val, "->", ifelse(srmr_val < 0.08, "Passed", "Failed"), "\n")
```

```{r, layout = c(1,1), results=F, message = F}
plot_network_graph (Usim_Grace_fit$beta, var.number = 9)

usemplot(Usim_Grace_fit$beta, var.number = 9, 
          labels = colnames(Grace_data_centered))
```

### Analysis of multiple individuals with uSEM

```{r, cache = TRUE}
df |> 
  group_by(name) |> 
  reframe(across(all_of(Vars), ~ nearZeroVar(as.data.frame(.))))
```

```{r, cache = TRUE}
# Initialize a list to store the results
results_list <- list()

# Define a vector of individual names (replace with actual names in your data)
individual_names <- unique(df$name)

# Loop through each individual dataset
for (individual in individual_names) {
  
  # Filter the dataset for the current individual
  individual_data <- df |> 
    filter(name == individual) |>
    select(all_of(Vars))
  
  # Center and scale the data
  centered_data <- individual_data |>
    mutate(across(everything(), ~ scale(.x, center = TRUE, scale = TRUE)))
  
  # Perform uSEM operations
  uSEM_result <- uSEM(var.number = 9, 
                      data = centered_data, 
                      lag.order = 1, 
                      verbose = FALSE,
                      trim = TRUE)
  
  # Get the model summary
  model_fit <- model_summary(model.fit = uSEM_result, 
                             var.number = 9, 
                             lag.order = 1)
  
  
  # Store the results in the list
  results_list[[individual]] <- list(
    uSEM_result = uSEM_result,
    model_fit = model_fit
  )
}
```

```{r, eval = F}
for (i in 20:24) {
  usemplot(results_list[[i]]$model_fit$beta, var.number = 9, labels = Vars,
           title = individual_names[i])
}
```
