---
title: "Explainable Artificial Intelligence in Education: A Tutorial for Identifying the Variables that Matter"
author: 
   - name: "Mohammed Saqr"
     email: "mohammed.saqr@uef.fi"
     affil-id: 1,*
   - name: "Sonsoles López-Pernas"
     email: "sonsoles.lopez@uef.fi"
     affil-id: 1
affiliations:
  - id: 1
    name: "University of Eastern Finland"
  - id: "*"
    name: "Corresponding author: Mohammed Saqr, `mohammed.saqr@uef.fi`"
---

# A tutorial on global xAI explainability using `DALEX`

## `DALEX` for student grade prediction 


```{r, setup, message = F}
set.seed(50)
# Load necessary libraries
library(tidyverse)
library(rsample)
library(e1071) 
library(DALEX)
library(rio)

# Import the data
student_data <- import("https://github.com/lamethods/data2/raw/main/lms/lms.csv")

# Standardize the numeric columns
student_data_standardized <- student_data |>
  mutate(across(where(is.numeric), ~scale(.) |> as.vector()))

# Split the data into training and testing sets
data_split <- initial_split(student_data_standardized, prop = 0.8)
train_data <- training(data_split)
test_data <- testing(data_split)

# Define the formula to specify the relationship between the target variable 
# and the predictor variables
formula <- Final_Grade ~ Freq_Course_View + Freq_Lecture_View + 
  Freq_Forum_Consume + Freq_Forum_Contribute + Regularity_Course_View +
  Regularity_Lecture_View + Regularity_Forum_Consume + 
  Regularity_Forum_Contribute + Session_Count + Total_Duration + Active_Days

# Fit the SVM model
svm_fit <- svm(formula,
               data = train_data,
               kernel = "radial")
```



```{r , warning=FALSE, message=FALSE}
set.seed(50)

library(DALEX)
# Create an explainer with DALEX
explainer_svm <- explain(
  svm_fit,
  data = test_data |> dplyr::select(-Final_Grade),
  y = test_data$Final_Grade,
  label = "SVM",
  verbose = FALSE
)
print(explainer_svm)
```


```{r, warning=FALSE, message=FALSE}
# Alternative code chunk
set.seed(50)

#Alternative model estimation for SVM

library(tidymodels)
# Define SVM model specification with a radial basis function kernel
svm_specification <- svm_rbf() |>
  set_mode("regression") |>
  set_engine("kernlab")

# Create the workflow. Combine the model specification with the formula
svm_workflow <- workflow() |>
  add_model(svm_specification) |>
  add_formula(formula)

# Fit the SVM model
svm_fit_tidy <- svm_workflow |>
  parsnip::fit(data = train_data)

# Create an explainer with DALEX
explainer_svm_tidy <- explain(
  svm_fit_tidy,
  data = test_data |> dplyr::select(-Final_Grade),
  y = test_data$Final_Grade,
  label = "SVM",
  verbose = FALSE)

# print(explainer_svm_tidy) # Uncomment to see the output
```


### Model evaluation


```{r, message=FALSE, warning=FALSE}
set.seed(50)

# Calculate and print model performance metrics for the SVM model
model_performance_svm <- model_performance(explainer_svm)
print(model_performance_svm)
```


```{r, message=FALSE, warning=FALSE, layout = matrix(rep(1,6),ncol = 2,byrow = T)}
set.seed(50)

# Generate various model performance plots
plot(model_performance_svm, geom = "lift") 
plot(model_performance_svm, geom = "gain") 
plot(model_performance_svm, geom = "boxplot") 
plot(model_performance_svm, geom = "histogram")
plot(model_performance_svm, geom = "ecdf")
```

### Model explainability

```{r, message=FALSE, warning=FALSE,  layout = matrix(rep(1,4) ,ncol = 2,byrow = T)}
set.seed(50)

# Default model
vi_svm <- model_parts(explainer_svm)
plot(vi_svm)

# Same as before with all arguments specified 
vi_svm <- model_parts(explainer_svm, loss_function = loss_root_mean_square, 
                      B = 100)
plot(vi_svm)  

# other variations: difference and ratio 
vi_svm_difference <- model_parts(explainer_svm, loss_function = loss_root_mean_square, 
                                 B = 100, type = "difference")
plot(vi_svm_difference)  

vi_svm_ratio <- model_parts(explainer_svm, loss_function = loss_root_mean_square, 
                            B = 100, type = "ratio")
plot(vi_svm_ratio)
```

#### Creating Custom Loss Functions for DALEX


```{r, message=FALSE, warning=FALSE, layout = c(1,1)}
set.seed(50)

# Load necessary libraries
library(DALEX)
library(ggplot2)

# Define the custom MAE loss function
loss_mae <- function(observed, predicted) {
  # Compute the Mean Absolute Error
  mean(abs(observed - predicted))
}

# Define the custom MAD loss function
loss_mad <- function(observed, predicted) {
  # Compute the Median Absolute Deviation
  median(abs(observed - predicted))
  #mean(abs(predicted - median(predicted))) another version
}

# Compute variable importance using the custom MAE loss function
vi_svm_mae <- model_parts(explainer_svm, loss_function = loss_mae)

# Compute variable importance using the custom MAD loss function
vi_svm_mad <- model_parts(explainer_svm, loss_function = loss_mad)

# Plot the results for MAE
plot(vi_svm_mae) 

# Plot the results for MAD
plot(vi_svm_mad)
```

### Model selection: creating and evaluating several ML models


```{r, message=FALSE, warning=FALSE}
set.seed(50)

# Load necessary packages
library(tidyverse)
library(caret)
library(rpart)
library(randomForest)
library(e1071)
library(kknn)
library(gbm)
library(xgboost)
library(DALEX)

# Linear Regression
linear_model <- lm(formula, data = train_data)

# Decision Tree
decision_tree_model <- rpart(formula, data = train_data, method = "anova")

# Random Forest
random_forest_model <- randomForest(formula, data = train_data, ntree = 100)

# Support Vector Machine
svm_model <- svm(formula, data = train_data)

# k-Nearest Neighbors
knn_model <- train(formula, data = train_data, method = "kknn", tuneLength = 5)

# Gradient Boosting Machine
gbm_model <- gbm(formula, data = train_data, distribution = "gaussian", 
                 n.trees = 100, interaction.depth = 3, shrinkage = 0.1, cv.folds = 5)

# XGBoost
train_matrix <- model.matrix(formula, data = train_data)[,c(-1)]
test_matrix <- model.matrix(formula, data = test_data)[,c(-1)]

train_label <- train_data$Final_Grade
xgboost_model <- xgboost(data = train_matrix, label = train_label, nrounds = 100, 
                         objective = "reg:squarederror", verbose = 0)
```

### Multiple models
#### Creating multiple explainers


```{r, message=FALSE, warning=FALSE}
set.seed(50)

# Create `DALEX`explainers
explainer_linear <- explain(
  model = linear_model, y = test_data$Final_Grade,
  data = dplyr::select(test_data, -Final_Grade),
  label = "Linear Regression", verbose = FALSE
)

explainer_decision_tree <- explain(
  model = decision_tree_model, y = test_data$Final_Grade,
  data = dplyr::select(test_data, -Final_Grade),
  label = "Decision Tree", verbose = FALSE
)

explainer_random_forest <- explain(
  model = random_forest_model, y = test_data$Final_Grade,
  data = dplyr::select(test_data, -Final_Grade),
  label = "Random Forest", verbose = FALSE
)

explainer_svm <- explain(
  model = svm_model, y = test_data$Final_Grade,
  data = dplyr::select(test_data, -Final_Grade),
  label = "SVM", verbose = FALSE
)

explainer_knn <- explain(
  model = knn_model$finalModel, y = test_data$Final_Grade,
  data = dplyr::select(test_data, -Final_Grade),
  label = "k-NN", verbose = FALSE
)

explainer_gbm <- explain(
  model = gbm_model, y = test_data$Final_Grade,
  data = dplyr::select(test_data, -Final_Grade),
  label = "GBM", verbose = FALSE
)

explainer_xgboost <- explain(
  model = xgboost_model, y = test_data$Final_Grade,
  data = test_matrix, label = "XGBoost", verbose = FALSE
)

```

#### Evaluation of multiple models


```{r, message=FALSE, warning=FALSE, results = FALSE}
set.seed(50)

# Calculate performance metrics for each model
performance_linear <- model_performance(explainer_linear)
performance_decision_tree <- model_performance(explainer_decision_tree)
performance_random_forest <- model_performance(explainer_random_forest)
performance_svm <- model_performance(explainer_svm)
performance_knn <- model_performance(explainer_knn)
performance_gbm <- model_performance(explainer_gbm)
performance_xgboost <- model_performance(explainer_xgboost)

# Combine performance metrics into a single table
performance_metrics <- rbind(
  `Linear Regression` = as.array(performance_linear$measure),
  `Decision Tree` = performance_decision_tree$measures, 
  `Random Forest` = performance_random_forest$measures,
  `SVM` = performance_svm$measures$r2, 
  `KNN` = performance_knn$measures,
  `GBM` = performance_gbm$measures,
  `XGBoost` = performance_xgboost$measures)  

# Print the combined dataframe
print(performance_metrics)
```



```{r, echo = FALSE}
plot.model_performance_histogram = function (df, nlabels) 
{
    diff <- label <- NULL
    ggplot(df, aes(diff, fill = label)) + geom_histogram(bins = 100) + 
        facet_wrap(~label, ncol = 2) + theme_default_dalex() + 
        xlab("residuals") + theme(legend.position = "none") + 
        scale_fill_manual(name = "Model", values = colors_discrete_drwhy(nlabels)) + 
        ggtitle("Histogram for residuals")
}

rlang::env_unlock(env = asNamespace('DALEX'))
rlang::env_binding_unlock(env = asNamespace('DALEX'))
assign('plot.model_performance_histogram', plot.model_performance_histogram, envir = asNamespace('DALEX'))
rlang::env_binding_lock(env = asNamespace('DALEX'))
rlang::env_lock(asNamespace('DALEX'))
```

```{r, message=FALSE, warning=FALSE, layout = matrix(c(rep(1,4),2,0),ncol = 2,byrow = T), fig.width=8, fig.height=5.4}
set.seed(50)

# Create combined plots
# Lift plot
plot(performance_linear, performance_decision_tree, performance_random_forest,
     performance_svm, performance_knn, performance_gbm, performance_xgboost,
    geom = "lift") 

# Gain plot
plot(performance_linear, performance_decision_tree, performance_random_forest,
     performance_svm, performance_knn, performance_gbm, performance_xgboost,
     geom = "gain") 

# Residuals plot
plot(performance_linear, performance_decision_tree, performance_random_forest,
     performance_svm, performance_knn, performance_gbm, performance_xgboost,
     geom = "boxplot"
)

# ecdf curve
plot(performance_linear, performance_decision_tree, performance_random_forest,
     performance_svm, performance_knn, performance_gbm, performance_xgboost,
     geom = "ecdf"
)

# histogram 
plot(performance_linear, performance_decision_tree, performance_random_forest,
     performance_svm, performance_knn, performance_gbm, performance_xgboost,
     geom = "histogram" 
) 
```


#### Explaining the XGBoost model


```{r, message=FALSE, warning=FALSE, layout = c(1,1,1)}
#| label: fig-xgboost-maemad
#| fig-cap: "XGBoost Variable importance using custom loss function"
#| fig-subcap: 
#|  - "RMSE"
#|  - "MAD"
#|  - "MAE"
set.seed(50)

# Compute and plot variable importance using the custom RMSE loss function
vi_xgboost <- model_parts(explainer_xgboost, loss_function = loss_root_mean_square,
                          B=500, type = "ratio")
plot(vi_xgboost)  

# Compute and plot variable importance using the custom MAD loss function
vi_xgboost_mad <- model_parts(explainer_xgboost, loss_function = loss_mad,  B=500)
plot(vi_xgboost_mad)

# Compute and plot variable importance using the custom MAE loss function
vi_xgboost_mae <- model_parts(explainer_xgboost, loss_function = loss_mae,B=500)
plot(vi_xgboost_mae) 
```

### Partial-dependence Profiles


#### Partial Dependence Plots (PDPs)


#### Accumulated Local Effects (ALE) Plots


```{r fig.width=8, fig.height=7, message=FALSE, warning=FALSE, layout = c(1,1)}
# Partial Dependence Plots
pdp_xgboost <- model_profile(explainer_xgboost, 
                variables = colnames(test_data)[colnames(test_data) != "Final_Grade"])
plot(pdp_xgboost) + labs(subtitle = "")

# Accumulated Local Effects (ALE) Plots
ale_xgboost <- model_profile(explainer_xgboost, type = "accumulated", 
                variables = colnames(test_data)[colnames(test_data) != "Final_Grade"])
plot(ale_xgboost) + labs(subtitle = "")
```


```{r fig.width=12, fig.height=6, message=FALSE, warning=FALSE}
# Accumulated Local Effects (ALE) Plots
ale_xgboost <- model_profile(explainer_xgboost, type = "accumulated", 
                variables = colnames(test_data)[colnames(test_data) != "Final_Grade"], 
                k = 3)
plot(ale_xgboost) + labs(subtitle = "")
```

## Explaining a classification model


```{r, message=FALSE, warning=FALSE}
# Load necessary libraries
library(gbm)
library(rio)
library(dplyr)
library(caret)
library(DALEX)
library(rsample)

# Set seed for reproducibility
set.seed(50)

# Load data
student_data <- import("https://github.com/lamethods/data2/raw/main/lms/lms.csv")

# Create binary classification target variable
student_data <- student_data |>
  mutate(Achievement = ifelse(Final_Grade > median(Final_Grade), 
                              "High_Achievers", "Low_Achievers")) |>
  mutate(Achievement = factor(Achievement, 
                              levels = c("High_Achievers", "Low_Achievers")))

# Convert target to numeric for model fitting
student_data$Achievement_numeric <- as.numeric(student_data$Achievement) - 1

# Split data into training and testing sets
data_split <- initial_split(student_data, prop = 0.8, strata = Achievement_numeric)
train_data <- training(data_split)
test_data <- testing(data_split)

# Fit GBM model
gbm_model <- gbm(
  Achievement_numeric ~  Freq_Course_View + Freq_Lecture_View + Freq_Forum_Consume + 
    Freq_Forum_Contribute + Regularity_Course_View + Regularity_Lecture_View + 
    Regularity_Forum_Consume + Regularity_Forum_Contribute +
    Session_Count + Total_Duration + Active_Days,
  data = train_data,
  distribution = "bernoulli",  # For binary classification
  n.trees = 1000,
  interaction.depth = 3,
  shrinkage = 0.01,
  bag.fraction = 0.7
)
```


```{r, message=FALSE, warning=FALSE}
# Create `DALEX` explainer
explainer_gbm <- explain(
  model = gbm_model,
  data = test_data[, 1:11],
  y = test_data$Achievement_numeric,
  label = "GBM Model"
)

# Model performance
model_performance_gbm <- model_performance(explainer_gbm)
print(model_performance_gbm)
```


```{r, message=FALSE, warning=FALSE, fig.width=5, fig.height=3, layout = matrix(rep(1,8), byrow = T, ncol = 2)}
# Create individual plots
plot(model_performance_gbm, geom = "roc")
plot(model_performance_gbm, geom = "boxplot")
plot(model_performance_gbm, geom = "lift")
plot(model_performance_gbm, geom = "gain")
plot(model_performance_gbm, geom = "prc")
plot(model_performance_gbm, geom = "histogram")
plot(model_performance_gbm, geom = "ecdf")
```


```{r, message=FALSE, warning=FALSE}
# Variable importance with AUC loss
vi_gbm <- model_parts(explainer_gbm, B = 500)
plot(vi_gbm)
```


```{r, layout = c(1,1,1), message=FALSE, warning=FALSE, results = FALSE}
# Custom loss functions (same as in the previous code)}
loss_logloss <- function(observed, predicted) {
  -mean(observed * log(predicted) + (1 - observed) * log(1 - predicted))
}

loss_f1_score <- function(observed, predicted) {
  predicted_class <- ifelse(predicted > 0.5, 1, 0)
  TP <- sum(observed == 1 & predicted_class == 1)
  FP <- sum(observed == 0 & predicted_class == 1)
  FN <- sum(observed == 1 & predicted_class == 0)
  F1 <- 2 * TP / (2 * TP + FP + FN)
  return(1 - F1)  # return 1 - F1 to keep lower values better
}

# Variable importance with custom loss functions
vi_gbm_logloss <- model_parts(explainer_gbm, loss_function = loss_logloss, 
                              B = 500, type = "ratio")
plot(vi_gbm_logloss)

vi_gbm_f1 <- model_parts(explainer_gbm, loss_function = loss_f1_score, B = 500)
plot(vi_gbm_f1)

vi_gbm_default <- model_parts(explainer_gbm, B = 500)
plot(vi_gbm_default)
```


```{r, message=FALSE, warning=FALSE, results = FALSE, layout = c(1,1), fig.width = 10, fig.height = 8}
# Partial Dependence Plots (PDP)
pdp_gbm <- model_profile(
  explainer_gbm,
  variables = colnames(test_data[, 1:11])
)
plot(pdp_gbm) + labs(subtitle = "")


# Accumulated Local Effects (ALE) Plots
ale_gbm <- model_profile(
  explainer_gbm,
  type = "accumulated",
  variables = colnames(test_data[, 1:11])
)
plot(ale_gbm) + labs(subtitle = "")
```
